Guilt, shame, and anti-war action in an authoritarian country at war

### 2024-06-12 manuscript.Rmd
Finished implementing revisions

d3$pcyn_ma <- rowMeans(d3[,c("pcyn1","pcyn2","pcyn3","pcyn4")],na.rm=TRUE)
describe(d3$pcyn_ma) #M=5.2,SD=1.5

d3$pcyn_im <- rowMeans(d3[,c("pcyn5","pcyn6","pcyn7","pcyn8")],na.rm=TRUE)
describe(d3$pcyn_im) #M=4.1,SD=1.52

d3$pal <- rowMeans(d3[,c("pal2","pal3","pal4")],na.rm=TRUE)
describe(d3$pal) #M=4.6,SD=1.38

d3$pr <- rowMeans(d3[,c("pr1","pr2","pr3","pr4","pr5")],na.rm=TRUE)
describe(d3$pr) #M=4.32,SD=1.15

@@

d3$posem <- rowMeans(d3[,c('happy','hope','enthusiasm','pride')], na.rm=TRUE)
describe(d3$posem) #M=2.19, SD=1.04

d3$neg_sub <- rowMeans(d3[,c('fear','sad','depressed')], na.rm=TRUE)
describe(d3$neg_sub) #M=2.88, SD=1.11

d3$neg_dom <- rowMeans(d3[,c('anger','disgust','contempt')], na.rm=TRUE)
describe(d3$neg_dom) #M=2.34, SD=1.25

@@

d3$cguilt <- rowMeans(d3[,c("cguilt1","cguilt2","cguilt3")], na.rm=TRUE)
describe(d3$cguilt) #N=400, M=2.97,SD=1.31
d3$cguiltf <- ifelse(is.na(d3$cguilt)&(d3$guilt==1),0,d3$cguilt)
describe(d3$cguiltf) #after imputation, M=1.23,SD=1.69

d3$mshame <- rowMeans(d3[,c("mshame1","mshame2","mshame3","mshame4")], na.rm=TRUE)
describe(d3$mshame) #N=349, M=3.02, SD=1.37
d3$mshamef <- ifelse(is.na(d3$mshame)&(d3$shame==1),0,d3$mshame)
describe(d3$mshamef) #after imputation, M=1.09,SD=1.67

d3$ishame <- rowMeans(d3[,c("ishame1","ishame2","ishame3","ishame4","ishame5")], na.rm=TRUE)
describe(d3$ishame) #N=349, M=3.13, SD=1.34
d3$ishamef <- ifelse(is.na(d3$ishame)&(d3$shame==1),0,d3$ishame)
describe(d3$ishamef) #after imputation, M=1.13,SD=1.7

@@

d3$bi_oppos <- rowMeans(d3[,c(179:181)],na.rm=TRUE)
describe(d3$bi_oppos) #M=1.53,SD=1.28
100*table(d3$bi_oppos)%>% prop.table()

@@

pdf('corplot.pdf', width=8, height=8)
plot <- d3[,c(216,218,220,170,251,182)]
colnames(plot) <- c('Guilt','Moral shame','Image shame', 'Past behavior',
                    'Behavioral intention', 'Action probability')

@@

*Support for democratic values* was measured with a 6-item scale from @CanettiNisim2007. Example item: "Every citizen has the right to take his convictions to the street if necessary" ($\alpha$ = `r round(alpha_demv$total[1],digits=2)`; M=`r `, SD=`r `).

@@

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = `r round(summary5$fit[9], digits = 3)`, RMSEA = `r round(summary5$fit[17], digits = 3)`, SRMR = `r round(summary5$fit[25], digits = 3)`). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = `r round(summary6$fit[9], digits = 3)`, RMSEA = `r round(summary6$fit[17], digits = 3)`, SRMR = `r round(summary6$fit[25], digits = 3)`). This modified model includes the *political responsibility* factor with 5 items (M=`r`, SD=`r`), *political alienation* factor with 3 items (M=`r`, SD=`r`; the item “I believe public officials don't care much what people like me think” was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: “*politicians are bad*” (4 items, e.g., “Politicians are only interested in getting and maintaining power”; M=`r`, SD=`r`) and “*politics is bad*” (4 items, e.g., “No man can hope to stay honest once he enters politics”; M=`r`, SD=`r`).

***Positive and negative emotions.*** We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale [PANAS; @Watson1988]. Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-point scale from 1 (*do not experience at all*) to 5 (*extremely*). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt (n=) and/or shame (n=), respectively. We took this precaution participants in earlier studies conducted in Russia complained about the double-barreled nature of items measuring group-based guilt and shame, where they felt that the questions imply that participants experience some level of guilt or shame (see @Grigoryan2018; @Grigoryan2017, for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame^[We excluded guilt and shame since we planned to estimate whether group-based guilt and shame are predictive of behavior over and above other emotions. Including PANAS measures of guilt and shame into the index of negative emotions would have confounded the two measures and prevented us from addressing this question (RQ1).]) using CFA and found the fit to be unsatisfactory (CFI = `r round(summary7$fit[9], digits = 3)`, RMSEA = `r round(summary7$fit[17], digits = 3)`, SRMR = `r round(summary7$fit[25], digits = 3)`). A 3-factor structure with *positive emotions* (joy, hope, enthusiasm; M=, SD=), *negative dominant emotions* (anger, disgust, contempt; M=, SD=) and *negative submissive emotions* (fear, sadness, depression; M=, SD=) showed a satisfactory fit to the data (CFI = `r round(summary8$fit[9], digits = 3)`, RMSEA = `r round(summary8$fit[17], digits = 3)`, SRMR = `r round(summary8$fit[25], digits = 3)`; see @Mehrabian1980 for the distinction between dominant and submissive emotions).

***Group-based guilt and shame.*** We used the measures of group-based guilt, image shame, and moral shame developed by @Rees2013, Russian adaptation by @Grigoryan2017. All items were answered on a 5-point scale from 1 (*completely disagree*) to 5 (*completely agree*). Guilt was measured with 3 items (e.g., “I feel guilty for the manner in which Ukrainian people have been treated by Russians”, $\alpha$ = `r round(alpha_guilt$total[1],digits=2)`), image shame with 5 items (“I feel ashamed when I realize that other countries might think of Russia negatively because of our involvement in Ukraine”, $\alpha$ = `r round(alpha_ishame$total[1],digits=2)`), and moral shame with 4 items (“I feel ashamed because Russia’s actions with regard to Ukraine have been immoral”, $\alpha$ = `r round(alpha_mshame$total[1],digits=2)`). All missing values were replaced with zeros, since participants who didn’t see the scale were the ones who reported no feelings of guilt or shame in the PANAS. With this imputation, we were able to use the whole dataset without excluding those who reported experiencing no guilt or shame. Mean values after this imputation were M=1.23 (SD=1.69) for guilt, M=1.13 (SD=1.70) for image shame, and M=1.09 (SD=1.67) for moral shame.

@@

***Anti-war actions***. We measured *past behavior* by asking how many of the eight political actions listed the participant had done to oppose the war in Ukraine (e.g., “signing a petition”, “participating in a demonstration”, etc.). The responses ranged from 0 to 8, with `r round(table(d3$pb_smop_opp)[1]/length(d3$pb_smop_opp)*100, digits = 0)`% of participants reporting zero actions, M=, SD=. *Behavioral intention* to participate in any political action against the war was measured with 3 items (“I want to/intend to/plan to take part in political action against the war in Ukraine”), on a 5-point scale from 1 (*completely disagree*) to 5 (*completely agree*) ($\alpha$ = `r round(alpha_bi_oppos$total[1],digits=2)`; `r round(table(means_bi)[1]/length(d3$bi_smop_opp1)*100, digits = 0)`% of participants had a mean score of 1; M=, SD=). Finally, we asked about *the probability that the person will take any political action*, from 1 (*highly improbable*), 10 (*(almost) certainly*) (`r round(table(d3$bp_smop_opp)[1]/length(d3$bp_smop_opp)*100, digits = 0)`% answered ‘1’; M=, SD=).

@@

###Descriptives (SM, Table S1)
apa.cor.table(
  d3[c("pb_smop_opp","bi_oppos", "bp_smop_opp",
       "mshamef","ishamef","cguiltf",
       "posem","neg_dom","neg_sub","smop_att",
       "demv","pcyn_ma","pcyn_im","pal","pr")],
  filename = "Cortable.doc",
  table.number = NA,
  show.conf.interval = TRUE,
  show.sig.stars = TRUE,
  landscape = TRUE
)

@@

h1ms <- lm(mshamef~demv+pcyn_ma+pcyn_im+pal+pr,d3)
summary(h1ms)
car::vif(h1ms)
h1ms2 <- lm(scale(mshamef)~scale(demv)+scale(pcyn_ma)+scale(pcyn_im)+scale(pal)+scale(pr),d3)
summary(h1ms)
confint(h1ms)

h1is <- lm(ishamef~demv+pcyn_ma+pcyn_im+pal+pr,d3)
summary(h1is)
car::vif(h1is)
h1is2 <- lm(scale(ishamef)~scale(demv)+scale(pcyn_ma)+scale(pcyn_im)+scale(pal)+scale(pr),d3)
summary(h1is)
confint(h1is)

h1g <- lm(cguiltf~demv+pcyn_ma+pcyn_im+pal+pr,d3)
summary(h1g)
car::vif(h1g)
h1g2 <- lm(scale(cguiltf)~scale(demv)+scale(pcyn_ma)+scale(pcyn_im)+scale(pal)+scale(pr),d3)
summary(h1g)
confint(h1g)

@@

##past behavior: linear (Table 2)
table(d3$pb_smop_opp)
h3a <- lm(pb_smop_opp ~ cguiltf+ishamef+mshamef,data = d3)
summary(h3a)
h3a2 <- lm(scale(pb_smop_opp) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef),data = d3)
summary(h3a2)
confint(h3a2)
car::vif(h3a2) #high VIF

h3guilt <- lm(scale(pb_smop_opp) ~ scale(cguiltf),data = d3)
summary(h3guilt) #.36***, r2=.13
h3ishame <- lm(scale(pb_smop_opp) ~ scale(ishamef),data = d3)
summary(h3ishame) #.39***, r2=.15
h3mshame <- lm(scale(pb_smop_opp) ~ scale(mshamef),data = d3)
summary(h3mshame) #.42***, r2=.18

@@

##behavioral intention (3 items): linear (Table 2)
table(d3$bi_oppos)
h3b <- lm(bi_oppos ~ cguiltf+ishamef+mshamef,data = d3)
summary(h3b)
h3b2 <- lm(scale(bi_oppos) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef),data = d3)
summary(h3b2)
confint(h3b2)
car::vif(h3b2) #high VIF

h3guilt <- lm(scale(bi_oppos) ~ scale(cguiltf),data = d3)
summary(h3guilt) #.55***, r2=.31
h3ishame <- lm(scale(bi_oppos) ~ scale(ishamef),data = d3)
summary(h3ishame) #.62***, r2=.39
h3mshame <- lm(scale(bi_oppos) ~ scale(mshamef),data = d3)
summary(h3mshame) #.65***, r2=.42

@@

##probability of doing something in the future: linear (Table 2)
table(d3$bp_smop_opp)
h3c <- lm(bp_smop_opp ~ cguiltf+ishamef+mshamef, data = d3)
h3coutput <- summary(h3c)
h3c2 <- lm(scale(bp_smop_opp) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef), data = d3)
summary(h3c2)
confint(h3c2)
car::vif(h3c2) #high VIF

h3guilt <- lm(scale(bp_smop_opp) ~ scale(cguiltf),data = d3)
summary(h3guilt) #.53***, r2=.28
h3ishame <- lm(scale(bp_smop_opp) ~ scale(ishamef),data = d3)
summary(h3ishame) #.57***, r2=.33
h3mshame <- lm(scale(bp_smop_opp) ~ scale(mshamef),data = d3)
summary(h3mshame) #.61***, r2=.37

@@

options("scipen"=100, "digits"=4)
h42 <- lm(fb_oppos ~ scale(posem)+scale(neg_dom)+scale(neg_sub)+
            scale(smop_att)+scale(mshamef)+scale(ishamef)+scale(cguiltf),d3)
summary(h42)

@@

h4b <- lm(fb_oppos ~ scale(posem)+scale(neg_dom)+scale(neg_sub)+
            scale(smop_att)+scale(mshamef)+scale(ishamef)+scale(cguiltf),d3)
summary(h4b)
confint(h4b)
car::vif(h4b)

h4r <- lm(fb_oppos ~ scale(posem)+scale(neg_dom)+scale(neg_sub)+
           scale(smop_att)+scale(ishamef)+scale(cguiltf),d3)
summary(h4r)
car::vif(h4r)

h4r2 <- lm(fb_oppos ~ scale(mshamef),d3)
summary(h4r2) #43%

h4r <- lm(fb_oppos ~ scale(cguiltf),d3)
summary(h4r)
#posem=14%,neg_dom=36%, neg_sub=17%, att=29%, ishame=38%, guilt=32%

@@

##exploratory analysis: predicting future anti-war action from anger, shame, and guilt in PANAS
panas <- lm(fb_oppos ~ anger+guilt+shame,d3)
summary(panas)
car::vif(panas)
panas <- lm(scale(fb_oppos) ~ scale(anger)+scale(guilt)+scale(shame),d3)
summary(panas)

panase <- lm(scale(fb_oppos) ~ scale(anger)+scale(mshamef)+scale(ishamef)+
             scale(cguiltf),d3)
summary(panase)

##exploratory analysis: interactions between political alienation and guilt and shame
int_guilt <- lm(scale(fb_oppos) ~ scale(cguiltf)*scale(pal),d3)
summary(int_guilt)

int_ms <- lm(scale(fb_oppos) ~ scale(mshamef)*scale(pal),d3)
summary(int_ms)

@@

knitr::include_graphics('corplot.pdf')

@@

#create matrix to contain the numbers of table 1
table1_numbers <- data.frame(matrix(nrow=7, ncol=12))
#add beta ms
table1_numbers[1:6,1] <- round(h1ms2$coefficients, digits = 2)
#add b ms
table1_numbers[1:6,2] <- round(h1ms$coefficients, digits = 2)
#add se ms
table1_numbers[1:6,3] <- round(summary_tablems$coefficients[1:6,2], digits = 2)
#add p values ms
table1_numbers[1:6,4] <- p_format(summary_tablems$coefficients[1:6,4], leading.zero = FALSE,digits=3,trailing.zero=TRUE,accuracy=0.001)

#add beta is
table1_numbers[1:6,5] <- round(h1is2$coefficients, digits = 2)
#add b is
table1_numbers[1:6,6] <- round(h1is$coefficients, digits = 2)
#add se is
table1_numbers[1:6,7] <- round(summary_tableis$coefficients[1:6,2], digits = 2)
#add p values is
table1_numbers[1:6,8] <- p_format(summary_tableis$coefficients[1:6,4], leading.zero = FALSE,digits=3,trailing.zero=TRUE,accuracy=0.001)

#add beta g
table1_numbers[1:6,9] <- round(h1g2$coefficients, digits = 2)
#add b g
table1_numbers[1:6,10] <- round(h1g$coefficients, digits = 2)
#add se g
table1_numbers[1:6,11] <- round(summary_tableg$coefficients[1:6,2], digits = 2)
#add p values g
table1_numbers[1:6,12] <- p_format(summary_tableg$coefficients[1:6,4], leading.zero = FALSE,digits=2,trailing.zero=TRUE,accuracy=0.001)

table1_numbers[7,c(2,6,10)] <- rms_table1
table1_numbers[7,c(1,3,4,5,7,8,9,11,12)] <- ' '

rownames(table1_numbers) <- c("Intercept", "Democratic Values", "Political Cynicism (politicians)", "Political Cynicism (politics)", "Political Alienation", "Political Responsibility","R$^{2}$")

colnames(table1_numbers) <- c("$\\beta$", "b", "SE", "p", "$\\beta$", "b", "SE", "p", "$\\beta$", "b", "SE", "p")

>>>>

d3$pcyn_ma <- rowMeans(d3[,c("pcyn1","pcyn2","pcyn3","pcyn4")],na.rm=TRUE)
pcyn_maout <- describe(d3$pcyn_ma) #M=5.2,SD=1.5

d3$pcyn_im <- rowMeans(d3[,c("pcyn5","pcyn6","pcyn7","pcyn8")],na.rm=TRUE)
pcyn_imout <- describe(d3$pcyn_im) #M=4.1,SD=1.52

d3$pal <- rowMeans(d3[,c("pal2","pal3","pal4")],na.rm=TRUE)
palout <- describe(d3$pal) #M=4.6,SD=1.38

d3$pr <- rowMeans(d3[,c("pr1","pr2","pr3","pr4","pr5")],na.rm=TRUE)
prout <- describe(d3$pr) #M=4.32,SD=1.15

@@

d3$posem <- rowMeans(d3[,c('happy','hope','enthusiasm','pride')], na.rm=TRUE)
posemout <- describe(d3$posem) #M=2.19, SD=1.04

d3$neg_sub <- rowMeans(d3[,c('fear','sad','depressed')], na.rm=TRUE)
neg_subout <- describe(d3$neg_sub) #M=2.88, SD=1.11

d3$neg_dom <- rowMeans(d3[,c('anger','disgust','contempt')], na.rm=TRUE)
neg_domout <- describe(d3$neg_dom) #M=2.34, SD=1.25

@@

d3$cguilt <- rowMeans(d3[,c("cguilt1","cguilt2","cguilt3")], na.rm=TRUE)
describe(d3$cguilt) #N=400, M=2.97,SD=1.31
d3$cguiltf <- ifelse(is.na(d3$cguilt)&(d3$guilt==1),0,d3$cguilt)
cguiltout <- describe(d3$cguiltf) #after imputation, M=1.23,SD=1.69

d3$mshame <- rowMeans(d3[,c("mshame1","mshame2","mshame3","mshame4")], na.rm=TRUE)
describe(d3$mshame) #N=349, M=3.02, SD=1.37
d3$mshamef <- ifelse(is.na(d3$mshame)&(d3$shame==1),0,d3$mshame)
mshameout <- describe(d3$mshamef) #after imputation, M=1.09,SD=1.67

d3$ishame <- rowMeans(d3[,c("ishame1","ishame2","ishame3","ishame4","ishame5")], na.rm=TRUE)
describe(d3$ishame) #N=349, M=3.13, SD=1.34
d3$ishamef <- ifelse(is.na(d3$ishame)&(d3$shame==1),0,d3$ishame)
ishameout <- describe(d3$ishamef) #after imputation, M=1.13,SD=1.7

@@

d3$bi_oppos <- rowMeans(d3[,c(179:181)],na.rm=TRUE)
bi_opposout <- describe(d3$bi_oppos) #M=1.53,SD=1.28
100*table(d3$bi_oppos)%>% prop.table()

@@

pdf('outputs/corplot.pdf', width=8, height=8)
plot <- d3[,c(216,218,220,170,251,182)]
colnames(plot) <- c('Guilt','Moral shame','Image shame', 'Past behavior',
                    'Behavioral intention', 'Action probability')

@@

*Support for democratic values* was measured with a 6-item scale from @CanettiNisim2007. Example item: "Every citizen has the right to take his convictions to the street if necessary" ($\alpha$ = `r round(alpha_demv$total[1],digits=2)`; M=`r round(alpha_demv$total[7],digits=2)`, SD=`r round(alpha_demv$total[8],digits=2)`).

@@

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = `r round(summary5$fit[9], digits = 3)`, RMSEA = `r round(summary5$fit[17], digits = 3)`, SRMR = `r round(summary5$fit[25], digits = 3)`). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = `r round(summary6$fit[9], digits = 3)`, RMSEA = `r round(summary6$fit[17], digits = 3)`, SRMR = `r round(summary6$fit[25], digits = 3)`). This modified model includes the *political responsibility* factor with 5 items (M=`r round(prout$mean, digits = 2)`, SD=`r round(prout$sd, digits = 2)`), *political alienation* factor with 3 items (M=`r round(palout$mean, digits = 2)`, SD=`r round(palout$sd, digits = 2)`; the item “I believe public officials don't care much what people like me think” was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: “*politicians are bad*” (4 items, e.g., “Politicians are only interested in getting and maintaining power”; M=`r round(pcyn_maout$mean, digits = 2)`, SD=`r round(pcyn_maout$sd, digits = 2)`) and “*politics is bad*” (4 items, e.g., “No man can hope to stay honest once he enters politics”; M=`r round(pcyn_imout$mean, digits = 2)`, SD=`r round(pcyn_imout$sd, digits = 2)`).

***Positive and negative emotions.*** We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale [PANAS; @Watson1988]. Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-point scale from 1 (*do not experience at all*) to 5 (*extremely*). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt (*n* = 400) and/or shame (*n* = 349), respectively. We took this precaution participants in earlier studies conducted in Russia complained about the double-barreled nature of items measuring group-based guilt and shame, where they felt that the questions imply that participants experience some level of guilt or shame (see @Grigoryan2018; @Grigoryan2017, for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame^[We excluded guilt and shame since we planned to estimate whether group-based guilt and shame are predictive of behavior over and above other emotions. Including PANAS measures of guilt and shame into the index of negative emotions would have confounded the two measures and prevented us from addressing this question (RQ1).]) using CFA and found the fit to be unsatisfactory (CFI = `r round(summary7$fit[9], digits = 3)`, RMSEA = `r round(summary7$fit[17], digits = 3)`, SRMR = `r round(summary7$fit[25], digits = 3)`). A 3-factor structure with *positive emotions* (joy, hope, enthusiasm; M=`r round(posemout$mean, digits = 2)`, SD=`r round(posemout$sd, digits = 2)`), *negative dominant emotions* (anger, disgust, contempt; M=`r round(neg_domout$mean, digits = 2)`, SD=`r round(neg_domout$sd, digits = 2)`) and *negative submissive emotions* (fear, sadness, depression; M=`r round(neg_subout$mean, digits = 2)`, SD=`r round(neg_subout$sd, digits = 2)`) showed a satisfactory fit to the data (CFI = `r round(summary8$fit[9], digits = 3)`, RMSEA = `r round(summary8$fit[17], digits = 3)`, SRMR = `r round(summary8$fit[25], digits = 3)`; see @Mehrabian1980 for the distinction between dominant and submissive emotions).

***Group-based guilt and shame.*** We used the measures of group-based guilt, image shame, and moral shame developed by @Rees2013, Russian adaptation by @Grigoryan2017. All items were answered on a 5-point scale from 1 (*completely disagree*) to 5 (*completely agree*). Guilt was measured with 3 items (e.g., “I feel guilty for the manner in which Ukrainian people have been treated by Russians”, $\alpha$ = `r round(alpha_guilt$total[1],digits=2)`), image shame with 5 items (“I feel ashamed when I realize that other countries might think of Russia negatively because of our involvement in Ukraine”, $\alpha$ = `r round(alpha_ishame$total[1],digits=2)`), and moral shame with 4 items (“I feel ashamed because Russia’s actions with regard to Ukraine have been immoral”, $\alpha$ = `r round(alpha_mshame$total[1],digits=2)`). All missing values were replaced with zeros, since participants who didn’t see the scale were the ones who reported no feelings of guilt or shame in the PANAS. With this imputation, we were able to use the whole dataset without excluding those who reported experiencing no guilt or shame. Mean values after this imputation were M=`r round(cguiltout$mean,digits=2)` (SD=`r round(cguiltout$sd,digits=2)`) for guilt, M=`r round(ishameout$mean,digits=2)` (SD=`r round(ishameout$sd,digits=2)`) for image shame, and M=`r round(mshameout$mean,digits=2)` (SD=`r round(mshameout$sd,digits=2)`) for moral shame.

@@

***Anti-war actions***. We measured *past behavior* by asking how many of the eight political actions listed the participant had done to oppose the war in Ukraine (e.g., “signing a petition”, “participating in a demonstration”, etc.). The responses ranged from 0 to 8, with `r round(table(d3$pb_smop_opp)[1]/length(d3$pb_smop_opp)*100, digits = 0)`% of participants reporting zero actions, M=`r round(mean(d3$pb_smop_opp,na.rm=TRUE),digits=2)`, SD=`r round(sd(d3$pb_smop_opp,na.rm=TRUE),digits=2)`. *Behavioral intention* to participate in any political action against the war was measured with 3 items (“I want to/intend to/plan to take part in political action against the war in Ukraine”), on a 5-point scale from 1 (*completely disagree*) to 5 (*completely agree*) ($\alpha$ = `r round(alpha_bi_oppos$total[1],digits=2)`; `r round(table(means_bi)[1]/length(d3$bi_smop_opp1)*100, digits = 0)`% of participants had a mean score of 1; M=`r round(bi_opposout$mean,digits=2)`, SD=`r round(bi_opposout$sd,digits=2)`). Finally, we asked about *the probability that the person will take any political action*, from 1 (*highly improbable*), 10 (*(almost) certainly*) (`r round(table(d3$bp_smop_opp)[1]/length(d3$bp_smop_opp)*100, digits = 0)`% answered ‘1’; M=`r round(mean(d3$bp_smop_opp,na.rm=TRUE),digits=2)`, SD=`r round(sd(d3$bp_smop_opp,na.rm=TRUE),digits=2)`).

@@

###Descriptives (SM, Table S1)
apa.cor.table(
  d3[c("pb_smop_opp","bi_oppos", "bp_smop_opp",
       "mshamef","ishamef","cguiltf",
       "posem","neg_dom","neg_sub","smop_att",
       "demv","pcyn_ma","pcyn_im","pal","pr")],
  filename = "outputs/Cortable.doc",
  table.number = NA,
  show.conf.interval = TRUE,
  show.sig.stars = TRUE,
  landscape = TRUE
)

msvims <- d3[c('mshamef','ishamef')]
msvimscor <- cor(msvims[which(!is.na(rowSums(msvims))),])[1,2]
msvguilt <- d3[c('mshamef','cguiltf')]
msvguiltcor <- cor(msvguilt[which(!is.na(rowSums(msvguilt))),])[1,2]

@@

h1ms <- lm(mshamef~demv+pcyn_ma+pcyn_im+pal+pr,d3)
summary(h1ms)
car::vif(h1ms)
h1ms2 <- lm(scale(mshamef)~scale(demv)+scale(pcyn_ma)+scale(pcyn_im)+scale(pal)+scale(pr),d3)
summary(h1ms)
h1ms_conf <- confint(h1ms2)

h1is <- lm(ishamef~demv+pcyn_ma+pcyn_im+pal+pr,d3)
summary(h1is)
car::vif(h1is)
h1is2 <- lm(scale(ishamef)~scale(demv)+scale(pcyn_ma)+scale(pcyn_im)+scale(pal)+scale(pr),d3)
summary(h1is)
h1is_conf <- confint(h1is2)

h1g <- lm(cguiltf~demv+pcyn_ma+pcyn_im+pal+pr,d3)
summary(h1g)
car::vif(h1g)
h1g2 <- lm(scale(cguiltf)~scale(demv)+scale(pcyn_ma)+scale(pcyn_im)+scale(pal)+scale(pr),d3)
summary(h1g)
h1g_conf <- confint(h1g2)

@@

##past behavior: linear (Table 2)
table(d3$pb_smop_opp)
h3a <- lm(pb_smop_opp ~ cguiltf+ishamef+mshamef,data = d3)
summary(h3a)
h3a2 <- lm(scale(pb_smop_opp) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef),data = d3)
summary(h3a2)
h3aCI <- confint(h3a2)
car::vif(h3a2) #high VIF

h3guilt <- lm(scale(pb_smop_opp) ~ scale(cguiltf),data = d3)
PB1 <- summary(h3guilt) #.36***, r2=.13
h3ishame <- lm(scale(pb_smop_opp) ~ scale(ishamef),data = d3)
PB2 <- summary(h3ishame) #.39***, r2=.15
h3mshame <- lm(scale(pb_smop_opp) ~ scale(mshamef),data = d3)
PB3 <- summary(h3mshame) #.42***, r2=.18

@@

##behavioral intention (3 items): linear (Table 2)
table(d3$bi_oppos)
h3b <- lm(bi_oppos ~ cguiltf+ishamef+mshamef,data = d3)
summary(h3b)
h3b2 <- lm(scale(bi_oppos) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef),data = d3)
summary(h3b2)
h3bCI <- confint(h3b2)
car::vif(h3b2) #high VIF

h3guilt <- lm(scale(bi_oppos) ~ scale(cguiltf),data = d3)
BI1 <- summary(h3guilt) #.55***, r2=.31
h3ishame <- lm(scale(bi_oppos) ~ scale(ishamef),data = d3)
BI2 <- summary(h3ishame) #.62***, r2=.39
h3mshame <- lm(scale(bi_oppos) ~ scale(mshamef),data = d3)
BI3 <- summary(h3mshame) #.65***, r2=.42

@@

##probability of doing something in the future: linear (Table 2)
table(d3$bp_smop_opp)
h3c <- lm(bp_smop_opp ~ cguiltf+ishamef+mshamef, data = d3)
h3coutput <- summary(h3c)
h3c2 <- lm(scale(bp_smop_opp) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef), data = d3)
summary(h3c2)
h3cCI <- confint(h3c2)
car::vif(h3c2) #high VIF

h3guilt <- lm(scale(bp_smop_opp) ~ scale(cguiltf),data = d3)
AP1 <- summary(h3guilt) #.53***, r2=.28
h3ishame <- lm(scale(bp_smop_opp) ~ scale(ishamef),data = d3)
AP2 <- summary(h3ishame) #.57***, r2=.33
h3mshame <- lm(scale(bp_smop_opp) ~ scale(mshamef),data = d3)
AP3 <- summary(h3mshame) #.61***, r2=.37

@@

options("scipen"=100, "digits"=4)
h42 <- lm(fb_oppos ~ scale(posem)+scale(neg_dom)+scale(neg_sub)+
            scale(smop_att)+scale(mshamef)+scale(ishamef)+scale(cguiltf),d3)
summary(h42)
h4CI <- confint(h42)

@@

h4b <- lm(fb_oppos ~ scale(posem)+scale(neg_dom)+scale(neg_sub)+
            scale(smop_att)+scale(mshamef)+scale(ishamef)+scale(cguiltf),d3)
summary(h4b)
confint(h4b)
vifout <- car::vif(h4b)

h4r <- lm(fb_oppos ~ scale(posem)+scale(neg_dom)+scale(neg_sub)+
           scale(smop_att)+scale(ishamef)+scale(cguiltf),d3)
excludingshame <- summary(h4r)
car::vif(h4r)

h4r2 <- lm(fb_oppos ~ scale(mshamef),d3)
shameonly <- summary(h4r2) #43%

h4r <- lm(fb_oppos ~ scale(cguiltf),d3)
guiltonly <- summary(h4r)
#posem=14%,neg_dom=36%, neg_sub=17%, att=29%, ishame=38%, guilt=32%

@@

##exploratory analysis: predicting future anti-war action from anger, shame, and guilt in PANAS
panas <- lm(fb_oppos ~ anger+guilt+shame,d3)
summary(panas)
car::vif(panas)
panas <- lm(scale(fb_oppos) ~ scale(anger)+scale(guilt)+scale(shame),d3)
explorout1 <- summary(panas)

panase <- lm(scale(fb_oppos) ~ scale(anger)+scale(mshamef)+scale(ishamef)+
             scale(cguiltf),d3)
explorout2 <- summary(panase)

##exploratory analysis: interactions between political alienation and guilt and shame
int_guilt <- lm(scale(fb_oppos) ~ scale(cguiltf)*scale(pal),d3)
explguiltout <- summary(int_guilt)

int_ms <- lm(scale(fb_oppos) ~ scale(mshamef)*scale(pal),d3)
explshameout <- summary(int_ms)

@@

knitr::include_graphics('outputs/corplot.pdf')

@@

#create matrix to contain the numbers of table 1
table1_numbers <- data.frame(matrix(nrow=21, ncol=7))
#add beta ms
table1_numbers[2:7,2] <- round(h1ms2$coefficients, digits = 2)
#add b ms
table1_numbers[2:7,5] <- round(h1ms$coefficients, digits = 2)
#add se ms
table1_numbers[2:7,6] <- round(summary_tablems$coefficients[1:6,2], digits = 2)
#add p values ms
table1_numbers[2:7,7] <- p_format(summary_tablems$coefficients[1:6,4], leading.zero = FALSE,digits=3,trailing.zero=TRUE,accuracy=0.001)

table1_numbers[2:7,3] <- round(h1ms_conf[,1],digits=2)
table1_numbers[2:7,4] <- round(h1ms_conf[,2],digits=2)

#add beta is
table1_numbers[9:14,2] <- round(h1is2$coefficients, digits = 2)
#add b is
table1_numbers[9:14,5] <- round(h1is$coefficients, digits = 2)
#add se is
table1_numbers[9:14,6] <- round(summary_tableis$coefficients[1:6,2], digits = 2)
#add p values is
table1_numbers[9:14,7] <- p_format(summary_tableis$coefficients[1:6,4], leading.zero = FALSE,digits=3,trailing.zero=TRUE,accuracy=0.001)

table1_numbers[9:14,3] <- round(h1is_conf[,1],digits=2)
table1_numbers[9:14,4] <- round(h1is_conf[,2],digits=2)

#add beta g
table1_numbers[16:21,2] <- round(h1g2$coefficients, digits = 2)
#add b g
table1_numbers[16:21,5] <- round(h1g$coefficients, digits = 2)
#add se g
table1_numbers[16:21,6] <- round(summary_tableg$coefficients[1:6,2], digits = 2)
#add p values g
table1_numbers[16:21,7] <- p_format(summary_tableg$coefficients[1:6,4], leading.zero = FALSE,digits=2,trailing.zero=TRUE,accuracy=0.001)

table1_numbers[16:21,3] <- round(h1g_conf[,1],digits=2)
table1_numbers[16:21,4] <- round(h1g_conf[,2],digits=2)

# table1_numbers[7,c(2,6,10)] <- rms_table1
table1_numbers[c(1,8,15),] <- ' '
table1_numbers[c(2,9,16),1:4] <- ' '
